ODGI: understanding pangenome graphs

### 2021-10-05 manuscript/fig_metrics.tex
finalized metrics figure caption

\begin{figure*}[h!]
	\begin{subfigure}{\linewidth}
		\centering
		% include first image
		\includegraphics[width=1.0\linewidth]{fig/metrics/chr4.pan.HTTex1.gfa.multiqc_odgi_stats.png}  
		\caption{Put your sub-caption here}
		\label{fig:metrics-multiqc}
	\end{subfigure}

	\begin{subfigure}{.5\linewidth}
		\centering
		% include second image
		\includegraphics[width=1.0\linewidth]{fig/metrics/chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og.png}  
		\caption{Put your sub-caption here}
		\label{fig:sub-second}
	\end{subfigure}
	\begin{subfigure}{.5\linewidth}
	\centering
	% include second image
	\includegraphics[width=1.0\linewidth]{fig/metrics/chr4.HTT.chm13.depth.w1.bed.pdf}  
	\caption{Put your sub-caption here}
	\label{fig:sub-sixth}
	\end{subfigure}

	\begin{subfigure}{.5\linewidth}
		\centering
		% include third image
		\includegraphics[width=1.0\linewidth]{fig/metrics/chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og.w2.convert.png}  
		\caption{Put your sub-caption here}
		\label{fig:sub-third}
	\end{subfigure}
	\begin{subfigure}{.5\linewidth}
		\centering
		% include fourth image
		\includegraphics[width=1.0\linewidth]{fig/metrics/chr4.HTT.chm13.degree.w1.bed.pdf}  
		\caption{Put your sub-caption here}
		\label{fig:sub-fourth}
	\end{subfigure}
%	\includegraphics[width=\linewidth]{fig/metrics/chr4.pan.HTTex1.gfa.multiqc_odgi_stats.png}
	\caption{Various outputs of the ODGI metrics tools.}
	\label{fig:metrics}
\end{figure*}

>>>>

\begin{figure*}[h!]
	\begin{subfigure}{\linewidth}
		\caption{}
		\centering
		% include first image
		\includegraphics[width=1.0\linewidth]{fig/metrics/chr4.pan.HTTex1.gfa.multiqc_odgi_stats.png}  
		\label{fig:metrics-multiqc}
	\end{subfigure}

	\begin{subfigure}{.5\linewidth}
		\caption{}
		\centering
		% include second image
		\includegraphics[width=1.0\linewidth]{fig/metrics/chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og.png}  
		\label{fig:metrics-viz}
	\end{subfigure}
	\begin{subfigure}{.5\linewidth}
		\caption{}
		\centering
		% include second image
		\includegraphics[width=1.0\linewidth]{fig/metrics/chr4.HTT.chm13.depth.w1.bed.pdf}  
		\label{fig:metrics-depth}
	\end{subfigure}

	\begin{subfigure}{.5\linewidth}
		\caption{}
		\centering
		% include third image
		\includegraphics[width=1.0\linewidth]{fig/metrics/chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og.w2.convert.png}  
		\label{fig:metrics-bin}
	\end{subfigure}
	\begin{subfigure}{.5\linewidth}
		\caption{}
		\centering
		% include fourth image
		\includegraphics[width=1.0\linewidth]{fig/metrics/chr4.HTT.chm13.degree.w1.bed.pdf}  
		\label{fig:metrics-degree}
	\end{subfigure}
%	\includegraphics[width=\linewidth]{fig/metrics/chr4.pan.HTTex1.gfa.multiqc_odgi_stats.png}
	\caption{Features of a human pangenome graph of the exon 1 huntingtin gene (\textit{HTTexon1}): \textbf{(a)} Excerpt of vital statistics of the \textit{HTTexon1} graph displayed by MultiQC's ODGI module. The very high GC content of 73.0\% compared to a human genomic mean GC content of 40.9\% \cite{Piovesan2019} is in accordance with the literature (see for example \cite{Sathasivam2013, Neueder2017}). \textbf{(b)} 1D visualization of the \textit{HTTexon1} graph with a bin width of 1. \textbf{(c)} Per nucleotide node depth distribution of CHM13 in the \textit{HTTexon1} graph. The alternating depth around position 200 indicates polymorphic variation. \textbf{(d)} 1D visualization of the \textit{HTTexon1} graph with a bin width of 2. Compared to above, the graph's width is exactly one half, therefore small variation is not visible anymore and the resolution of graph topology decreases. \textbf{(e)} Per nucleotide node degree distribution of CHM13 in the \textit{HTTexon1} graph. The polymorphic region is around position 200 complementing the above node depth analysis. Figures \textbf{(b)-(e)} clearly highlight the variant region around position 200.}
	\label{fig:metrics}
\end{figure*}

### 2021-10-05 manuscript/main.tex
finalized metrics figure caption

Applying \texttt{odgi stats}, users can retrieve metrics describing the graph properties, such as the number of nodes, edges, paths, and graph length. MultiQC's \citep{Ewels_2016} ODGI module\footnote{\url{https://multiqc.info/docs/\#odgi} (accessed Oct 2021)} provides an interactive way to comparatively explore the statistics of an arbitrary number of graphs. ODGI also offers more advanced tools for the interrogation of the graphs. \texttt{odgi bin} summarizes the path information into bins of a specified size, enabling a summarized view of gigabase scale graphs. This concept found its way into the pangenome graph ontology \cite{Yokoyama2020}. \texttt{odgi depth} returns the node depth as defined by query criteria, allowing users to assess the complexity of regions in the graph. Collapsed sequences with highly identical repeats indicate intricate loci. Complex motifs can be also detected with \texttt{odgi degree}, which returns the node degree as defined by query criteria. High degree nodes are the mirror of misassemblies or problems in the pangenome building, making the tool useful for validation. In Fig. \ref{fig:metrics}, the characteristics of a human pangenome graph of the exon 1 huntingtin gene \citep{Sathasivam2013} (\textit{HTTexon1}), obtained with the ODGI metrics tools, are shown.

@@

rdf sparql side, bring in annotation for great biological interpretation

@@

(recent) interactive visualization tools?!

>>>>

Applying \texttt{odgi stats}, users can retrieve metrics describing the graph properties, such as the number of nodes, edges, paths, and graph length. MultiQC's \citep{Ewels_2016} ODGI module\footnote{\url{https://multiqc.info/docs/\#odgi} (accessed Oct 2021)} provides an interactive way to comparatively explore the statistics of an arbitrary number of graphs. ODGI also offers more advanced tools for the interrogation of the graphs. \texttt{odgi bin} summarizes the path information into bins of a specified size, enabling a summarized view of gigabase scale graphs. This concept found its way into the pangenome graph ontology \cite{Yokoyama2020}. \texttt{odgi depth} returns the node depth as defined by query criteria, allowing users to assess the complexity of regions in the graph. Collapsed sequences with highly identical repeats indicate intricate loci. Complex motifs can be also detected with \texttt{odgi degree}, which returns the node degree as defined by query criteria. High degree nodes are the mirror of misassemblies or problems in the pangenome building, making the tool useful for validation. In Fig. \ref{fig:metrics}, the characteristics of a human pangenome graph of the exon 1 huntingtin gene \citep{Sathasivam2013,Neueder2017} (\textit{HTTexon1}), obtained with the ODGI metrics tools, are shown.

@@

rdf sparql side, bring in annotation for great biological interpretation \cite{Yokoyama2020}, federated queries, sharing, FAIR

(recent) interactive visualization tools?! CITE PANACHE, also indicate that no published tool for interactive exploration of such graphs exist

@@
